### 2021-06-22 rad-icfp-2021/appendix1.tex
minors
changing makefile and adding todos
writing appendix

We show that the gradients from the introduction 
can be obtained as special instances of our general construction.

@@

@@

\begin{lemma}
    $\nabla_A$!dot(A,B)! := !map2 * B (map (x -> 1) A)! 
\end{lemma}

>>>>

 \subsection{Operational semantics}

In Figure~\ref{fig:op_semantics_target} a small step call-by-value operational semantics for the language. 
\MH{need to add evaluation contexts}

Some operators are less conventional which we now explain.
!shift1! takes an array of reals of size $n$ and adds a $1$ as first component and forgets about the last component of the array, returning an array of size !n!.
!scanr! is similar to !scanl! except that the folding starts from the end of the array. 
It is equivalent to first reversing the input array, performing !scanl!, and reversing the resulting array.

\MH{need to explain that the !scanl!, !scanr! coming from !reduce! are also parallel ones}

\MH{might need to extend with the operators from the generalized section}

\input{figures/op_semantics_target}

 \subsection{Reverse derivative of array operations}

We now prove that the reverse mode transformation is correct on array operations. 
The proofs consist in showing that unrolling the computation before differentiation 
and then differentiating gives the same as differentiation and then unrolling.

 \begin{proposition}
     The reverse derivative of !map! is correct.
 \end{proposition}

@@

 \begin{proposition}
    The reverse derivative of !map2! is correct.
\end{proposition}

\begin{proposition}
    The reverse derivative of !foldl! is correct.
\end{proposition}

We show that the gradients from Section \ref{sec:intro} are obtained as instances of our general construction. 
The proofs consist in instantiating the general derivatives to these cases 
and showing that each rewrite step is a simple known optimization.

Similarly to numpy, we use the notation OnesLike(A) to mean map (x -> 1) A 
and ZerosLike(A) to mean map (x -> 0) A. 
Finding these constant arrays is key to a lot of optimisations that leverage the ring algebraic structure of the reals to arrays.

\MH{change to make it nice. be more explicit about optimisations. 0 should be ZerosLike. 
Also, I am freely referring to the variables from the context or bound in theoriginal term, should be clarified.}

 \begin{proof}
\begin{align*}
    \nabla_A \text{prod(A)} 
    &= \nabla_A(\text{reduce} * 1 A) \\
    &\defeq  \Big(\lambda A',z.
    \begin{aligned}
        & let A_1 = scanl * 1 A in \\
        & let A_2 = map2 (x,y -> y) A_1 A in\\
        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in\\
        & let A_4 = scanr * 1 A_2 in \\
        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
    \end{aligned}
    \Big)(0,1)\\
    &\stackrel{\beta-reduction}{=}  \begin{aligned}
        & let A_1 = scanl * 1 A in &\\
        & let A_2 = map2 (x,y -> y) A_1 A in & A_2=  A\\
        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 A1\\
        & let A_4 = scanr * 1 A_2 in & \\
        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
    \end{aligned}\\
    &=  \begin{aligned}
        & let A_1 = scanl * 1 A in & \\
        & let A_3 = shift1 A1 in &\\
        & let A_4 = scanr * 1 A in &\text{forward substitution} A_2\\
        & map2 * A_4 A_3
    \end{aligned}\\
    &\stackrel{\eta-reduction}{=} map2 * (scanr x 1 A) (shift1(scanl * 1 A))
\end{align*}
 \end{proof}

@@

\begin{align*}
    \nabla_A sum(A) 
    &= \nabla_A reduce + 0 A \\
    & \defeq \Big(\lambda A',z.
    \begin{aligned}
        & let A_1 = scanl * 1 A in \\
        & let A_2 = map2 (x,y -> 1) A_1 A in \\
        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
        & let A_4 = scanr * 1 A_2 in \\
        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
    \end{aligned}
    \Big)(0,1)\\
    &{\beta-reduction}{=}
    \begin{aligned}
        & let A_1 = scanl * 1 A in &\\
        & let A_2 = map2 (x,y -> y) A_1 A in & A_2= OnesLike(A) \\
        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 OnesLike(A)\\
        & let A_4 = scanr * 1 A_2 in & \\
        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
    \end{aligned}\\
    &= \begin{aligned}
        & let A_1 = scanl * 1 A in & \\
        & let A_2= OnesLike(A) in &  \\
        & let A_3 = shift1 OnesLike(A) in & A_3 = OnesLike(A) \\
        & let A_4 = scanr * 1 A_2 in & A_4 = OnesLike(A_2) \\
        & map2 * A_4 A_3 &
    \end{aligned}\\
    &= map2 * OnesLike(A) OnesLike(A) \\
    &= OnesLike(A)
\end{align*}
 \end{proof}

 \begin{lemma}
     $\nabla_A$!dot(A,B)! = !B! 
 \end{lemma}

 \begin{proof}
The reverse derivative of !map2 * A B! is given by 
\begin{align*}
    & \lambda A',B',C'. \\
    & let C_1 =map2 (a,b -> b) A B in \\
    & let C_2 =map2 (a,b -> a) A B in \\
    & (map3 (a,b,c -> a+b*c) A' C_1 C', map3 (a,b,c -> a+b*c) A' C_2 C')
\end{align*}

Let's call this term Y.
For convenience, let us also rewrite !map2 * B (map (x -> 1) A)! 
as !let C = map (x -> 1) A in map2 * B C!.

Then the gradient of !dot(A,B)! is given by
\begin{align*}
    \nabla dot(A,B) 
    &= \nabla let C = map (x -> 1) A in map2 * B C\\
    &\defeq \lambda A',B',C',z. 
    \left.\begin{aligned}
        & let A_1 = scanl * 1 A in \\
        & let A_2 = map2 (x,y -> 1) A_1 A in \\
        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
        & let A_4 = scanr * 1 A_2 in \\
        & Y(A', B', map3 (a,b,c -> a+b*c*z) A' A_4 A_3)
    \end{aligned}\right\}same reduction as previously
    \\
    &\quad(0,0,0,1) \\
    &= Y(0,0,OnesLike(C)) \\
    &\defeq 
    \begin{aligned}
        & let C_1 =map2 (a,b -> b) A B in & C_1 = B\\
        & let C_2 =map2 (a,b -> a) A B in & C_2 = A\\
        & (map3 (a,b,c -> a+b*c) 0 C_1 OnesLike(C), & map2 & C_1 OnesLike(C)
        & map3 (a,b,c -> a+b*c) 0 C_2 OnesLike(C)) & map2 & C_2 OnesLike(C)
    \end{aligned}\\
    &= \begin{aligned}
        & let C_1 = B & \text{forward substitution}\\
        & let C_2 = A \text{forward substitution}\\
        & map2 & C_1 OnesLike(C) & C_1
        & map2 & C_2 OnesLike(C) & C_2
    \end{aligned}\\
    &= (B, A)
\end{align*}
So if we are only interested in the gradient w.r.t. A, this indeed gives us B.
 \end{proof}

### 2021-06-22 rad-icfp-2021/complexity.tex
outline paper in intro
writing appendix

\section{Complexity}

@@

>>>>

\section{Complexity}
\label{sec:complexity}

@@

We follow a simple model similar to the one in \cite{griewank2008evaluating}.
We assume the cost is divided into 4 elementary measures, being the number of MOVES, ADDS, MULTS, and NLOPS.
MOVES assumes a flat memory and represents moving a fixed size information  (e.g. 64 bits). 
ADDS represents the number of additions, 
MULTS the number of multiplications, 
and NLOPS the number of elementary non linear operations like !cos!, !exp!.

The gives a complexity function $\cost$ valued in $\RR^4$. 
For instance, we have

\begin{itemize}
    \item $\cost(*)=(3,0,1,0)$
    \item $\cost(+)=(3,1,0,0)$
    \item $\cost(c)=(1,0,0,0)$
    \item $\cost($!sin!$)=(2,0,0,1)$
\end{itemize}

\MH{need cost array operations}
\MH{need to extend the cost function in a compositional way}
